On the Minimum Core Mass for Giant Planet Formation at Wide Separations

### 2013-10-25 tex/draft/draft4.tex
minor section 4 edits + removed model breakdown part

\newcommand{\eq}[1]{equation\,(\ref{#1})}
\newcommand{\Eq}[1]{Equation\,(\ref{#1})}
\newcommand{\Eqs}[2]{Equations (\ref{#1}) and~(\ref{#2})}
\newcommand{\Eqss}[2]{Equations (\ref{#1})--(\ref{#2})}
\newcommand{\Eqsss}[3]{Equations (\ref{#1}), (\ref{#2}) and~(\ref{#3})}
\newcommand{\App}[1]{Appendix~\ref{#1}}
\newcommand{\Sec}[1]{Sect.~\ref{#1}}
\newcommand{\Chap}[1]{Chapter~\ref{#1}}
\newcommand{\Fig}[1]{Fig.~\ref{#1}}
\newcommand{\Figs}[2]{Figs.~\ref{#1} and \ref{#2}}
\newcommand{\Figss}[2]{Figs.~\ref{#1}--\ref{#2}} 
\newcommand{\Tab}[1]{Table \ref{#1}}

@@

\begin{figure}[tb]
\centering
%AY: figure should probably use R_B for consistency with text.
\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/PTm_profiles_v2_smallM.eps}
%\vspace{-0.5in}
\caption{Radial profiles of atmospheric pressure, temperature and enclosed mass (core included) for a $5 M_{\oplus}$ core at $a=60$ AU.   Solid, dotted, dashed and dot-dashed lines correspond to solutions with total mass (core and atmosphere) of $5.01 M_{\oplus}$, $5.10 M_{\oplus}$, $6.00 M_{\oplus}$ and $9.24 M_{\oplus}$, respectively.  (See text for a description of evolution to yet higher masses.)   Circles and triangles mark the locations of the Bondi radii and of the radiative-convective boundaries, respectively.  The radial profiles extend from the core to the Hill radius.} 
\label{fig:profiles}
\end{figure}

\subsection{Atmospheric Stucture}
\label{sec:profiles}
Figure \ref{fig:profiles} shows radial profiles at different stages of atmospheric growth around a $5 M_{\oplus}$ core at $60$ AU.  Quoted mass values include the core plus atmosphere within the smaller of $\RB$ or $\RH$, which for these cases is $\RB$.  The 9.24 $M_{\oplus}$ solution is the highest mass we can reach in our evolutionary sequence, as we explain further in \S\ref{sec:endoftime}.

The lowest mass atmosphere -- which we take as our initial state -- is fully convective with the same entropy as the disk.  In \Fig{fig:profiles} this state is the 5.01 $M_{\oplus}$ solution with no radiative zone.  Atmospheric cooling and contraction allows the accretion of more gas.  In the convective zones, higher mass solutions contain lower entropy and higher pressures.  A radiative zone emerges to connect the lower entropy interior to the higher entropy disk.  \Fig{fig:profiles} shows that this radiative zone is already fairly deep in the 5.10 $M_\oplus$ solution.

@@

The cooling model of \S\ref{sec:twolayer} is used to connect solutions of different atmospheric masses into an evolutionary sequence.  \Fig{fig:Ltplot} shows the luminosity evolution and the elapsed time as a function of atmospheric mass for a $5 M_{\oplus}$ core at $60$ AU, i.e.\ the same parameters as in \Fig{fig:profiles}.  During the early stages of atmospheric growth, the luminosity drops sharply.  This behavior is seen in both the full numerical solutions (with solid lines) and the analytic model (with dashed lines).  With increasing atmospheric mass, the pressure depth of the RCB increases, along with the optical depth ($\propto \kappa P\cb)$.  Consequently, the radiative luminosity decreases.  This behavior is described in \Eqs{eq:crit}{eq:Lcb}.

At later stages of evolution, the numerical model in \Fig{fig:Ltplot} shows a flat luminosity with increasing mass and also time (not shown).  By contrast, the non-self-gravitating analytic model gives a luminosity that continues to drop with increasing mass.  To understand this difference, consider the scaling of \Eq{eq:Lcb}, $L\cb \propto M\cb T\cb^4/ (\kappa\cb  P\cb)$, which holds in both cases.   Accounting for the higher enclosed mass in the self-gravitating model gives a somewhat higher luminosity, as desired.  However, the main effect is that \Eq{eq:crit} -- which describes a nearly linear relation between atmospheric mass and RCB pressure --  breaks down for self-gravitating solutions.  This behavior can be seen in the top panel of \Fig{fig:profiles} where the $P\cb$ increases significantly from 5.10 to 6.0 $M_\oplus$, but only increases relatively modestly with further growth to 9.24 $M_\oplus$.   In the higher mass solutions, the relatively low $P\cb$ values (and thus the relatively high luminosities) require an outward shift in $R\cb$, as shown in \Fig{fig:profiles}.

The accelerated growth in the numerical model, as shown in the bottom panel of \Fig{fig:Ltplot}, is also a direct result of the enhanced cooling (higher luminosities) in the self-gravitating models.  We see that self-gravity plays a very important role in atmospheric evolution, even well before the crossover mass.

@@

The fact that growth times have a well defined maximum is a characteristic of runaway growth.  Unlike our analytic model, which must assume that runaway growth begins near the crossover mass, our numerical model allows us to measure when runaway growth begins.  Runaway growth clearly does not begin at a universal value of $M_{\rm atm}/M\co$.  \Fig{fig:growthtime} shows that further from the star runaway growth begins at smaller $M_{\rm atm}/M\co$.  \Fig{fig:tvsM} (described in the next section) shows how the onset of runaway growth depends on core mass.

@@

At even higher masses than shown in \Fig{fig:coolingterms}, our cooling model breaks down.  Mathematically, the breakdown manifests as a negative time step (due to an increase in internal entropy) when advancing to a higher atmospheric mass.  This unphysical result has no deep significance, as the neglected luminosity overwhelms the model luminosity by this point.    Effectively the model is telling us when to ignore it.  Our results are unaffected by the breakdown, which occurs comfortably after $t_{\rm run}$, our threshold for runaway growth.

>>>>

\newcommand{\eq}[1]{equation\,(\ref{#1})}
\newcommand{\Eq}[1]{Equation\,(\ref{#1})}
\newcommand{\Eqs}[2]{Equations (\ref{#1}) and~(\ref{#2})}
\newcommand{\Eqss}[2]{Equations (\ref{#1})--(\ref{#2})}
\newcommand{\Eqsss}[3]{Equations (\ref{#1}), (\ref{#2}) and~(\ref{#3})}
\newcommand{\App}[1]{Appendix~\ref{#1}}
\newcommand{\Sec}[1]{Sect.~\ref{#1}}
\newcommand{\Chap}[1]{Chapter~\ref{#1}}
\newcommand{\Fig}[1]{Figure~\ref{#1}}
\newcommand{\Figs}[2]{Figs.~\ref{#1} and \ref{#2}}
\newcommand{\Figss}[2]{Figs.~\ref{#1}--\ref{#2}} 
\newcommand{\Tab}[1]{Table \ref{#1}}

@@

\begin{figure}[tb]
\centering
%AY: figure should probably use R_B for consistency with text.
\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/PTm_profiles_v2_smallM.eps}
%\vspace{-0.5in}
\caption{Radial profiles of atmospheric pressure, temperature and enclosed mass (core included) for a $5 M_{\oplus}$ core at $a=60$ AU.   Solid, dotted, dashed and dot-dashed lines correspond to solutions with total mass (core and atmosphere) of $5.01 M_{\oplus}$, $5.10 M_{\oplus}$, $6.00 M_{\oplus}$ and $8.99 M_{\oplus}$, respectively.  Circles and triangles mark the locations of the Bondi radii and of the radiative-convective boundaries, respectively.  The radial profiles extend from the core to the Hill radius.} %  (See text for a description of evolution to yet higher masses.) 
\label{fig:profiles}
\end{figure}

\subsection{Atmospheric Stucture}
\label{sec:profiles}
Figure \ref{fig:profiles} shows radial profiles at different stages of atmospheric growth around a $5 M_{\oplus}$ core at $60$ AU.  Quoted mass values include the core plus atmosphere within the smaller of $\RB$ or $\RH$, which for these cases is $\RB$.  The 8.99 $M_{\oplus}$ solution is the mass at which runaway growth begins according to the criteria described in \S\ref{sec:timeev}. % we can reach in our evolutionary sequence, as we explain further in \S\ref{sec:endoftime}.

The lowest mass atmosphere -- which we take as our initial state -- is fully convective with the same entropy as the disk.  In \Fig{fig:profiles} this state is the 5.01 $M_{\oplus}$ solution with no radiative zone.  Atmospheric cooling and contraction allows the accretion of more gas.  In the convective zones, higher mass solutions have lower entropy and higher pressures.  A radiative zone emerges to connect the lower entropy interior to the higher entropy disk.  \Fig{fig:profiles} shows that this radiative zone is already fairly deep in the 5.10 $M_\oplus$ solution.

@@

The cooling model of \S\ref{sec:twolayer} is used to connect solutions with different atmospheric masses into an evolutionary sequence.  \Fig{fig:Ltplot} shows the luminosity evolution and the elapsed time as a function of atmospheric mass for a $5 M_{\oplus}$ core at $60$ AU, i.e.\ the same parameters as in \Fig{fig:profiles}.  During the early stages of atmospheric growth, the luminosity drops sharply.  This behavior is seen in both the full numerical solutions (with solid lines) and the analytic model (with dashed lines).  With increasing atmospheric mass, the pressure depth of the RCB increases, along with the optical depth ($\propto \kappa P\cb)$.  Consequently, the radiative luminosity decreases.  This behavior is described in \Eqs{eq:crit}{eq:Lcb}.

At later stages of evolution, the numerical model in \Fig{fig:Ltplot} shows a flat luminosity with increasing mass and also time (not shown).  By contrast, the non-self-gravitating analytic model gives a luminosity that continues to drop as the atmosphere becomes more massive.  To understand this difference, consider the scaling of \Eq{eq:Lcb}, $L\cb \propto M\cb T\cb^4/ (\kappa\cb  P\cb)$, which holds in both cases.   Accounting for the higher enclosed mass in the self-gravitating model gives a somewhat higher luminosity, as desired.  However, the main effect is that \Eq{eq:crit} -- which describes a nearly linear relation between atmospheric mass and RCB pressure --  breaks down for self-gravitating solutions.  This behavior can be seen in the top panel of \Fig{fig:profiles} where the $P\cb$ increases significantly from 5.10 to 6.0 $M_\oplus$, but only increases relatively modestly with further growth to 9.24 $M_\oplus$.   In the higher mass solutions, the relatively low $P\cb$ values (and thus the relatively high luminosities) require an outward shift in $R\cb$, as shown in \Fig{fig:profiles}.

The accelerated growth in the numerical model, as shown in the bottom panel of \Fig{fig:Ltplot}, is also a direct result of the enhanced cooling (higher luminosities) in the self-gravitating models.  We see that self-gravity plays a very important role in atmospheric evolution, even well before the crossover mass, when $M_{\rm atm}=M\co$.

@@

The fact that growth times have a well defined maximum is a characteristic of accelerating growth.  Unlike our analytic model, which must assume that runaway growth begins near the crossover mass, our numerical model allows us to measure when runaway accretion begins.  Runaway growth clearly does not begin at a universal value of $M_{\rm atm}/M\co$.  \Fig{fig:growthtime} shows that further from the star runaway growth begins at smaller $M_{\rm atm}/M\co$.  \Fig{fig:tvsM} (described in the next section) shows how the onset of runaway growth depends on core mass.

@@

%At even higher masses than shown in \Fig{fig:coolingterms}, our cooling model breaks down.  Mathematically, the breakdown manifests as a negative time step (due to an increase in internal entropy) when advancing to a higher atmospheric mass.  This unphysical result has no deep significance, as the neglected luminosity overwhelms the model luminosity by this point.    Effectively the model is telling us when to ignore it.  Our results are unaffected by the breakdown, which occurs comfortably after $t_{\rm run}$, our threshold for runaway growth.
